Evolution of the AGN UV luminosity function from redshift 7.5

### 2018-04-06 Paper/paper.tex
Update text

%% Q notation is different on figure and in text.
%% MH15 curve seems wrong; it should reionize at z ~ 4 or so.
\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth,keepaspectratio]{q.pdf}
  \end{center}
  \caption{Evolution of the \HeIII\ ionization fraction in our model,
    when the AGN luminosity function is integrated down to
    $M_{1450}=-21$ (blue curve) and $M_{1450}=-18$ (red curve).  The
    shaded regions show the one-sigma (68.26\%) uncertainty.  Other
    curves show models of \citet[solid grey]{2012ApJ...746..125H},
    \citet[dashed grey]{2015ApJ...813L...8M}, \citet[light blue and
      shaded region]{2016ApJ...828...90L},
    \citet[blue]{2018arXiv180104931P},
    \citet[brown]{2009ApJ...694..842M}, and
    \citet[green]{2014MNRAS.445.4186C}.  Note that we derive
    $Q_\nHeIII$ using Equation~(\ref{eqn:qHedot}), which ignores the
    presence of \HeII\ Lyman-limit systems and becomes inaccurate
    towards the end of \HeII\ reionization. \label{fig:qhe}}
\end{figure}

We now consider the implications of our AGN luminosity function models
for \HeII\ reionization.  We use the emissivity model developed in
Section~\ref{sec:e912} for this purpose.  The evolution of the
volume-averaged fraction $Q_\nHeIII$ of \HeIII\ is given by
\citep{2012ApJ...746..125H}
\begin{equation}
{dQ_\nHeIII\over dt}={\dot n_{\rm ion,4}\over \langle n_\nHe \rangle} -{Q_\nHeIII\over
\langle t_{\rm rec,He}\rangle},
\label{eqn:qHedot}
\end{equation}
where $\dot n_{\rm ion,4}$ is the number density of photons with
energy 4~Ry and above produced per unit time, $n_\nHe$ is the Helium
number density, and $\langle t_{\rm rec,He}\rangle$ is an average
recombination time scale for \HeIII.  We obtain the evolution of $\dot
n_{\rm ion,4}$ for AGN with $M_\mathrm{1450}<-21$ and
$M_\mathrm{1450}<-18$ by integrating the contribution of double power
law luminosity function obtained in distinct redshifts bins and
fitting Equation~(\ref{eqn:e912fit}) to the resultant \HeII-ionizing
emissivities.  We extrapolate the resultant fit to $z>7$, although the
contribution from this extrapolated emissivity to \HeII-reionization
is small.  As before, we assume the SEDs given by
Equations~(\ref{eqn:sed}) and (\ref{eqn:sed_faint}) for bright and
faint AGN, respectively.  The recombination time scale in
Equation~(\ref{eqn:qHedot}) is given by
\begin{equation}
\langle t_{\rm rec}\rangle=[(1+2\chi) \langle n_\nH\rangle \alpha_B\,C]^{-1},
\label{eq:trec}
\end{equation}
where $n_\nH$ is the average physical number density of hydrogen,
$\alpha_B$ is the case-B \HeIII\ recombination rate, and $\chi=0.083$
is the cosmic number fraction of Helium.  We assume that the helium
clumping factor $C$ is given by \citep{2015ApJ...813L...8M}
\begin{equation}
  C = 2.9\left(\frac{1+z}{6}\right)^{-1.1}.
\end{equation}
We use the fitting function provided by \citet{1997MNRAS.292...27H}
for the recombination coefficient.

Note that Equation~(\ref{eqn:qHedot}) neglects the \HeII\ Lyman-limit
systems that play an increasingly important role towards the end of
Helium reionization \citep{2009MNRAS.395..736B, 2017ApJ...851...50M}.
In the absence of these self-shielded systems of high-density \HeII,
$Q_\nHeIII$ can continue to increase beyond unity.  We set it to one
when this happens.  This can be corrected by accounting for the
\HeII\ column density distribution and filtering the \HeII-ionizing
radiation background through it.  Unfortunately, the \HeII\ column
density distribution is itself uncertain.  We therefore consider our
treatment of $Q_\nHeIII$ accurate enough for the purpose of this work.

The resultant reionization histories of \HeII\ are shown in
Figure~\ref{fig:qhe} for $M_\mathrm{1450}<-18$ and
$M_\mathrm{1450}<-21$ in red and blue, respectively.  We find that
Helium reionization is complete, respectively, at $z\sim 3.2$ and
$3.1$ in these models, with a one-sigma uncertainty of about $\Delta
z\sim 0.2$.  Observational constraints on the \HeII\ reionization
history are derived from measurements of the effective \HeII\ \lya
optical depth \citep{2001Sci...293.1112K, 2004ApJ...600..570S,
  2006A&A...455...91F, 2010ApJ...722.1312S, 2011ApJ...733L..24W,
  2016ApJ...825..144W}.  While we do not compute the effective
\HeII\ \lya optical depth in our model, the reionization histories
shown in Figure~\ref{fig:qhe} may likely be consistent with the
optical depth measurements of \citet{2016ApJ...825..144W}, who find
that the effective optical depth rises at $z>2.7$, but low optical
depth sightlines continue to be seen beyond this redshift.  The
incidence of these low opacity sightlines implies that the substantial
volume of \HeII\ was ionized already at $z=3.4$.  \HeII\ reionization
begins at $z\gtrsim 5$ in our models.  When we consider AGN with
$M_\mathrm{1450}<-18$, the ionization fraction $Q_\nHeIII$ has a
one-sigma spread of 0.8--1 at $z=3.4$.

In Figure~\ref{fig:qhe}, we also show the reionization histories from
other models of \HeII\ reionization in the literature.  Our model
curves are closest in agreement to the simulations of
\citet{2009ApJ...694..842M} and the analytical model of
\citet{2018arXiv180104931P}.  \citet{2018arXiv180104931P} model the
\HeII-ionizing emissivity of AGN by fitting the function form in
Equation~(\ref{eqn:e912fit}) to the emissivities reported by
\citet{2009A&A...507..781S}, \citet{2007A&A...472..443B}, and
\citet{2012ApJ...755..169M}.  The resultant emissivity in the model of
\citet{2018arXiv180104931P} is similar in shape but higher in
amplitude than the emissivity in the \citet{2012ApJ...746..125H}
model.  The \HeII-ionizing emissivity of \citet{2018arXiv180104931P}
is closer to our inference for $M_\mathrm{1450}<-21$ than the model of
\citet{2012ApJ...746..125H}.  (However, note that both these models
suffer from inhomogeneous luminosity function integration limits, as
the integration limit is set to a fixed fraction of the break
luminosity (cf.\ discussion in Section~\ref{sec:e912}.)  A spectral
index of $\alpha_\mathrm{UV}=-1.7$ is assumed.
\citet{2018arXiv180104931P} compute the evolution of $Q_\nHeIII$ by
evolving the \HeIII\ ionization fraction in a gas cloud with the mean
cosmic baryonic density exposed to the average \HeII-ionizing flux.
The ionizing flux is obtained by filtering the ionizing emissivity
through an inhomogeneous IGM in which the column density distribution
of \HeIII\ is derived from that of \HI\ in a manner similar to
\citet{2012ApJ...746..125H}.  As a result, this model incorporates the
effect of \HeII\ Lyman-limit systems, and the $Q_\nHeIII$ evolution in
this model deviates from our predictions towards the end of
reionization, as seen in Figure~\ref{fig:qhe}.  Comparing the
emissivity model in the cosmological radiative transfer simulations of
\citet{2009ApJ...694..842M} with our models is less straightforward.
Quasars are modelled in the simulations of \citet{2009ApJ...694..842M}
by relating quasar luminosities and lifetimes to halo masses following
the galaxy-merger-induced quasar activity model of
\citet{2005ApJ...630..705H}.  The resultant bolometric quasar
luminosity functions are somewhat shallower than those of
\citet{2007ApJ...654..731H}, but with a higher amplitude.  A spectral
index of $\alpha_\mathrm{UV}=-1.6$ is assumed.  Figure~\ref{fig:qhe}
shows that the resultant reionization history in these simulations
agrees with our model for $M_\mathrm{1450}<-21$, except towards the
end of reionization, where, as before, \HeII\ Lyman-limit systems
become important.

Helium reionization is delayed in the model of
\citet{2012ApJ...746..125H} due to their reduced emissivity at $z\sim
3$.  In the \citet{2015ApJ...813L...8M} model, a significant
\HeII\ fraction is ionized already at $z=6$ due to the very high
emissivity assumed in this model.  In the radiation-hydrodynamic
simulations by \citet{2016ApJ...828...90L}, the AGN population is
calibrated to the SDSS and COSMOS data at $z>3.5$
\citep{2012ApJ...755..169M, 2013ApJ...768..105M}, combined with a
model for quasar light curves.  The fiducial \HeII\ reionization
history obtained by these authors is shown in Figure~\ref{fig:qhe}
along with an estimate of the uncertainty.  Reionization ends
considerably later in this model ($z\sim 2.6$).  Our redshift of
reionization agrees more closely with that in the simulations
presented by \citet{2014MNRAS.445.4186C}.

Helium reionization has also been studied by
\citet{2018MNRAS.473.1416M}, who constrained the evolution of the
emissivity from AGN using the effect optical depth measurement of
\citet{2016ApJ...825..144W}, combined with an analytical model of the
IGM.  While a detailed comparison with their work is difficult because
of reasons, \citet{2018MNRAS.473.1416M} find that the AGN emissivity
has to drop considerably rapidly at $z>4$. \citet{2018arXiv180109693K}
have recently presented a model of Helium reionization.  Reionization
ends at redshift 2.8 in their model.  The AGN luminosity function used
is by a combination of deteminations from various data sets
\citep{2015MNRAS.451L..30K}.  The integration limits are inconsistent
however as they are $0.01L_*$.  Resulting emissivity is comparable to
our model at $z\sim 2$ but drops more rapidly towards higher
redshifts.  For a steep EUV spectral index of $\alpha=-1.8$ their
helium reionization redshift is $z=2.8$.  \citet{2016arXiv160706467D}
considered the effect of AGN-dominated reionization on the Ly$\alpha$
opacity at $z>5$, He~\textsc{ii} Ly$\alpha$ opacity at $z\sim
3.1$--$3.3$, and the thermal history of the IGM.  In agreement with
\citet{2015MNRAS.453.2943C}, these authors found that AGN did provide
a plausible explanation for the large fluctuations in the Ly$\alpha$
opacity at $z>5$.  However, they found that reionization of
He~\textsc{ii} occurs much earlier in these AGN-dominated models (see
also \citealt{2018MNRAS.473.1416M}).

>>>>

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth,keepaspectratio]{q.pdf}
  \end{center}
  \caption{Evolution of the \HeIII\ ionization fraction in our model,
    when the AGN luminosity function is integrated down to
    $M_{1450}=-21$ (blue curve and shaded region) and $M_{1450}=-18$
    (red curve and shaded region).  The shaded regions show the
    one-sigma (68.26\%) uncertainty.  Other curves show models of
    \citet[solid grey]{2012ApJ...746..125H}, \citet[dashed
      grey]{2015ApJ...813L...8M}, \citet[brown with shaded
      region]{2016ApJ...828...90L}, \citet[blue]{2018arXiv180104931P},
    \citet[maroon]{2009ApJ...694..842M}, and
    \citet[green]{2014MNRAS.445.4186C}.  Note that we derive
    $Q_\nHeIII$ using Equation~(\ref{eqn:qHedot}), which ignores the
    presence of \HeII\ Lyman-limit systems and becomes inaccurate
    towards $Q_\nHeIII\sim 1$. \label{fig:qhe}}
\end{figure}

We now consider the implications of our AGN luminosity function models
for \HeII\ reionization.  We use the emissivity model developed in
Section~\ref{sec:e912} for this purpose.  The evolution of the
volume-averaged fraction $Q_\nHeIII$ of \HeIII\ is given by
\citep{2012ApJ...746..125H}
\begin{equation}
{dQ_\nHeIII\over dt}={\dot n_{\rm ion,4}\over \langle n_\nHe \rangle} -{Q_\nHeIII\over
\langle t_{\rm rec,He}\rangle},
\label{eqn:qHedot}
\end{equation}
where $\dot n_{\rm ion,4}$ is the number density of photons with
energy 4~Ry and above produced per unit time, $n_\nHe$ is the Helium
number density, and $\langle t_{\rm rec,He}\rangle$ is an average
recombination time scale for \HeIII.  We obtain the evolution of $\dot
n_{\rm ion,4}$ for AGN with $M_\mathrm{1450}<-21$ and
$M_\mathrm{1450}<-18$ by integrating the contribution of double power
law luminosity function obtained in distinct redshifts bins and
fitting Equation~(\ref{eqn:e912fit}) to the resultant \HeII-ionizing
emissivities.  We extrapolate the resultant fit to $z>7$, although the
contribution from this extrapolated emissivity to \HeII-reionization
is small.  As before, we assume the SED given by
Equations~(\ref{eqn:sed}), which has a spectral index of
$\alpha_\mathrm{EUV}=-1.7$.  The recombination time scale in
Equation~(\ref{eqn:qHedot}) is given by
\begin{equation}
\langle t_{\rm rec}\rangle=[(1+2\chi) \langle n_\nH\rangle \alpha_B\,C]^{-1},
\label{eq:trec}
\end{equation}
where $\langle n_\nH\rangle$ is the average physical number density of
hydrogen, $\alpha_B$ is the case-B \HeIII\ recombination rate, and
$\chi=0.079$ is the cosmic number fraction of helium for a cosmic
helium mass fraction of $Y_\mathrm{He}=0.24$.  The hydrogen number
density is given by $\langle n_\nH\rangle=1.881\times
10^{-7}(1+z)^3$~cm$^{-3}$.  We assume that the helium clumping factor
$C$ is given by \citep{2015ApJ...813L...8M}
\begin{equation}
  C = 2.9\left(\frac{1+z}{6}\right)^{-1.1}.
\end{equation}
We use the fitting function provided by \citet{1997MNRAS.292...27H}
for the recombination coefficient.

Equation~(\ref{eqn:qHedot}) neglects the \HeII\ Lyman-limit systems
that play an increasingly important role towards the end of Helium
reionization \citep{2009MNRAS.395..736B, 2017ApJ...851...50M}.  In the
absence of these self-shielded systems of high-density \HeII,
$Q_\nHeIII$ can continue to increase beyond unity.  The mean free path
of \HeII-ionizing photon diverges.  We set $Q_\nHeIII=1$ when this
happens.  This can be corrected by accounting for the \HeII\ column
density distribution and filtering the \HeII-ionizing radiation
background through it.  Unfortunately, the \HeII\ column density
distribution is itself uncertain \citep{2018arXiv180104931P}.  We
consider our treatment of $Q_\nHeIII$ accurate enough for the purpose
of this work, while noting that the redshift of \HeII\ reionization is
likely an overestimate in our model.

The resultant reionization histories of \HeII\ are shown in
Figure~\ref{fig:qhe} for $M_\mathrm{1450}<-18$ and
$M_\mathrm{1450}<-21$.  When only AGN with $M_\mathrm{1450}<-21$ are
accounted, Helium reionization is complete at $z\sim
3.20^{+0.08}_{-0.12}$ where the uncertainty is one-sigma.  When the
luminosity function is extrapolated down to $M_\mathrm{1450}<-18$,
\HeII\ is reionized at $z=3.47^{+0.22}_{-0.26}$.  \HeII\ reionization
begins at $z\gtrsim 5$ in both models.  Observational constraints on
the \HeII\ reionization history are derived from measurements of the
effective \HeII\ \lya optical depth \citep{2001Sci...293.1112K,
  2004ApJ...600..570S, 2006A&A...455...91F, 2010ApJ...722.1312S,
  2011ApJ...733L..24W, 2016ApJ...825..144W}.  These are considered to
imply a redshift of \HeII\ reionization of $z=2.7$
\citep{2011ApJ...733L..24W}.  While we do not compute the effective
\HeII\ \lya optical depth in our model, the higher reionization
redshifts in our models are partly explained by the absence of
\HeII\ Lyman-limit systems, as discussed above.  Furthermore, the
reionization histories shown in Figure~\ref{fig:qhe} may likely be
consistent with the optical depth measurements of
\citet{2016ApJ...825..144W}, who find that the effective optical depth
rises at $z>2.7$, but low optical depth sightlines continue to be seen
beyond this redshift.  The incidence of these low opacity sightlines
implies that the substantial volume of \HeII\ was ionized already at
$z=3.4$, consistent with our models.

In Figure~\ref{fig:qhe}, we also show the reionization histories from
other models of \HeII\ reionization in the literature.  Our model
curve for $M_\mathrm{1450}<-21$ is closest in agreement to the
simulations of \citet{2009ApJ...694..842M} and the analytical model of
\citet{2018arXiv180104931P}.  \citet{2018arXiv180104931P} model the
\HeII-ionizing emissivity of AGN by fitting the functional form in
Equation~(\ref{eqn:e912fit}) to the emissivities reported by
\citet{2009A&A...507..781S}, \citet{2007A&A...472..443B}, and
\citet{2012ApJ...755..169M}.  The resultant emissivity in the model of
\citet{2018arXiv180104931P} is similar in shape but higher in
amplitude than the emissivity in the \citet{2012ApJ...746..125H}
model.  The \HeII-ionizing emissivity of \citet{2018arXiv180104931P}
is closer to our inference for $M_\mathrm{1450}<-21$ than the model of
\citet{2012ApJ...746..125H}.  (However, note that both these models
suffer from inhomogeneous luminosity function integration limits, as
the integration limit is set to a fixed fraction of the break
luminosity, as discussed above in Section~\ref{sec:e912}.)  A spectral
index of $\alpha_\mathrm{EUV}=-1.7$ is assumed.
\citet{2018arXiv180104931P} compute the evolution of $Q_\nHeIII$ by
evolving the \HeIII\ ionization fraction in a gas cloud with the mean
cosmic baryonic density exposed to the average \HeII-ionizing flux.
The ionizing flux is obtained by filtering the ionizing emissivity
through an inhomogeneous IGM in which the column density distribution
of \HeIII\ is derived from that of \HI\ in a manner similar to
\citet{2012ApJ...746..125H}.  As a result, this model incorporates the
effect of \HeII\ Lyman-limit systems, and the $Q_\nHeIII$ evolution in
this model deviates from our predictions towards the end of
reionization, as seen in Figure~\ref{fig:qhe}.  Comparing the
emissivity model in the cosmological radiative transfer simulations of
\citet{2009ApJ...694..842M} with our models is less straightforward.
Quasars are modelled in the simulations of \citet{2009ApJ...694..842M}
by relating quasar luminosities and lifetimes to halo masses following
the galaxy-merger-induced quasar activity model of
\citet{2005ApJ...630..705H}.  The resultant bolometric quasar
luminosity functions are somewhat shallower than those of
\citet{2007ApJ...654..731H}, but with a higher amplitude.  A spectral
index of $\alpha_\mathrm{EUV}=-1.6$ is assumed.  Figure~\ref{fig:qhe}
shows that the resultant reionization history in these simulations
agrees with our model for $M_\mathrm{1450}<-21$, except towards the
end of reionization, where, as before, \HeII\ Lyman-limit systems
become important.

Helium reionization is delayed in the model of
\citet{2012ApJ...746..125H} due to their reduced emissivity at $z\sim
3$.  In the \citet{2015ApJ...813L...8M} model, a significant
\HeII\ fraction is ionized already at $z=6$, and reionization is
complete by $z=4.1$ due to the very high emissivity assumed in this
model.  In the radiation-hydrodynamic simulations by
\citet{2016ApJ...828...90L}, the AGN population is calibrated to the
SDSS \citep{2013ApJ...768..105M} and COSMOS
\citep{2012ApJ...755..169M} data at $z>3.5$, combined with a model for
quasar light curves.  The fiducial \HeII\ reionization history
obtained by these authors is shown in Figure~\ref{fig:qhe} along with
an estimate of the uncertainty.  Reionization ends considerably later
in this model ($z\sim 2.6$).  The \HeII\ reionization history in our
model for $M_\mathrm{1450}<-18$ agrees closely with that in the
radiative transfer simulations presented by
\citet{2014MNRAS.445.4186C}.  The AGN population in these simulations
is `turned on' at $z=5$ and evolved assuming a pure luminosity
evolution to match the inferred luminosity function of
\citet{2011ApJ...728L..26G}.  The somewhat early reionization in the
simulations of \citet{2014MNRAS.445.4186C} relative to most other
models shown in Figure~\ref{fig:qhe} may be driven by the
significantly higher emissivity from AGN with intermediate brightness
($M_{1450}\sim -23$) in the model of \citet{2011ApJ...728L..26G}.
However, the limited numerical resolution of these simulations may
also play a role \citep{2014MNRAS.445.4186C}.

An important constraint on \HeII\ reionization can be derived from the
thermal history of the IGM.  \citet{2018MNRAS.473.1416M} constrained
the evolution of the 912\,\AA\ emissivity of AGN by assuming a
spectral index of $\alpha_\mathrm{EUV}=-1.57$ and using the
measurements of the effective \HeII\ \lya optical depth
\citep{2016ApJ...825..144W} and the IGM temperature
\citep{2011MNRAS.410.1096B}.  \citet{2018MNRAS.473.1416M} found that
although these data favour a reionization history in which \HeII\ is
reionized at $z>4$, the 912\,\AA\ emissivity of AGN is nevertheless
required to drop rapidly at $z>3$.  This emissivity evolution is much
more rapid than that in our model for $M_\mathrm{1450}<-21$, with the
caveat that the constraint derived by \citet{2018MNRAS.473.1416M} is
model-dependent and assumes a shallower EUV spectrum.  A steeper
spectrum with $\alpha_\mathrm{EUV}=-1.8$ was assumed in a recent model
by \citet{2018arXiv180109693K}.  The resultant emissivity evolution is
consistent with our model for $M_\mathrm{1450}<-21$, although somewhat
more rapid at $z>2$.  The AGN luminosity function is derived by
combining determinations from various data sets
\citep{2015MNRAS.451L..30K}.  An inhomogeneous integration limit of
$0.01L_*$ is applied to calculate the emissivity.  \HeII\ reionization
ends at $z=2.8$ in the model of \citet{2018arXiv180109693K}.
(\citealt{2017MNRAS.471..255K} argued that for their luminosity
function model, spectra with $\alpha_\mathrm{EUV}>-1.6$ are unlikely
to be consistent with the \HeII\ \lya data.)
\citet{2017MNRAS.468.4691D} considered the effect of an AGN-dominated
reionization on the \HI\ Ly$\alpha$ opacity at $z>5$,
\HeII\ Ly$\alpha$ opacity at $z\sim 3.1$--$3.3$, and the thermal
history of the IGM.  These authors found that that any scenario in
which reionization of \HeII\ completes at $z>3$ is inconsistent with
the measurements of the IGM temperature at $z=4$--$5$.  We leave the
analysis of the IGM thermal history implied by our AGN luminosity
models for future work.
